Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-17 paper/intro.tex
review comments

Spectral distributions can reveal important properties in numerous applications across physics, chemistry, engineering, and data science: For example, in electronic structure calculations eigenvalues represent the energy levels which electrons occupy\cite{drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure, lin-2017-randomized-estimation}, in machine learning they are indicative of the topology of the loss landscape \cite{ghorbani-2019-investigation-neural, yao-2020-pyhessian-neural}, and in graph processing they can uncover hidden graph motifs~\cite{huang-2021-density-states}. Given a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, the distribution of its eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$, which places a rescaled Dirac delta distribution $\delta$ at each eigenvalue. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation that is often prohibitively expensive. Several techniques have been developed for approximating $\phi$, including moment matching~\cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time} and polynomial approximation, either implicitly by a Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or explicitly by expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}.

@@

For a \emph{constant} matrix $\mtx{B}$, one of the most popular trace estimators is the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} along with variance reduction techniques~\cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, chen-2023-krylovaware-stochastic, epperly-2024-xtrace-making}. Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing insight and justification of these extensions. In passing, we note that dynamic trace estimation~\cite{dharangutte-2024-dynamic-trace,woodruff-2024-optimal-query} is an efficient technique for subsequently estimating the traces of matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ for wich the increments $\mtx{B}(t_{i+1}) - \mtx{B}(t_i)$ are relatively small in norm. The potential of dynamic trace estimation appears to be limited in our setting because $\mtx{B}(t)$ may change rapidly close to eigenvalues, with $g_{\sigma}$ approximating a Dirac delta function.

>>>>

Spectral distributions of matrices and linear operators reveal important properties in numerous applications across physics, chemistry, engineering, and data science: For example, in electronic structure calculations, eigenvalues represent the energy levels that electrons occupy~\cite{drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure, lin-2017-randomized-estimation}, in machine learning they are indicative of the topology of the loss landscape~\cite{ghorbani-2019-investigation-neural, yao-2020-pyhessian-neural}, and in graph processing they can uncover hidden graph motifs~\cite{huang-2021-density-states}. Given a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, the distribution of its eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$, which places a rescaled Dirac delta distribution $\delta$ at each eigenvalue. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation that is often prohibitively expensive. Several techniques have been developed for approximating $\phi$, including moment matching~\cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time} and polynomial approximation, either implicitly by a Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or explicitly by expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Some of these methods have also been combined with a deflation step to yield even more efficient algorithms \cite{lin-2017-randomized-estimation, bhattacharjee-2024-improved-spectral}.

@@

For a \emph{constant} matrix $\mtx{B}$, one of the most popular trace estimators is the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} along with variance reduction techniques~\cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, chen-2023-krylovaware-stochastic, epperly-2024-xtrace-making}. Suitable extensions to parameter-dependent matrices have been considered, e.g., in~\cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but we are not aware of an analysis providing rigorous justification and insight of these extensions. In passing, we note that dynamic trace estimation~\cite{dharangutte-2024-dynamic-trace,woodruff-2024-optimal-query} is an efficient technique for subsequently estimating the traces of matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ when the increments $\mtx{B}(t_{i+1}) - \mtx{B}(t_i)$ are relatively small in norm. The potential of dynamic trace estimation appears to be limited in our setting because $\mtx{B}(t)$ may change rapidly close to eigenvalues, with $g_{\sigma}$ approximating a Dirac delta function.
